---
title: "Storm Database analysis: most devastating events in the United States, years 1952 - 2011"
author: Andrey Budish
date: 2017-02-20
output:
  html_notebook:
    highlight: tango
    theme: united
    toc: yes
---



## 1. Synopsis
Analysis showed the 5 most devastating health related events were: tornado (96979 total casualties), excessive heat (8428), tsunami wind (7461), flood (7259) and lightning (6046).  
Specifically, the top 5 most devastating events related to injuries: tornado, tsunami wind, flood, excessive heat and lightning;
the top 5 most damaging related to fatalities: tornado, excessive heat, flash flood, heat and lightning.  

As for top 5 events related to the greatest economic consequences, those were: tornado (3312277), flash flood (1599325), tsunami wind (1445168), hail (1268290) and flood (1067976).  
Specifically, the top 5 most damaging related to property: tornado, flash flood, tsunami wind, flood and thunderstorm wind;
the top 5 most devastating related to crops: hail, flash flood, flood, tsunami wind and tornado.


## 2. Data Processing  

#### Set global options for knitr:
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=6)
```


#### Display when and on what system analysis was performed:  
```{r date}
dateDownloaded <- date()
dateDownloaded
```

```{r system_info}
system_info <- sessionInfo()
system_info
```

#### Download storm dataset and archive:  
```{r download}
dlMethod <- "curl" # sets default for OSX / Linux
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(substr(Sys.getenv("OS"),1,7) == "Windows") dlMethod <- "wininet"

if(!file.exists("StormData.csv.bz2")) {
  download.file(url,
                "StormData.csv.bz2",  # stores file in R working directory
                method=dlMethod # use OS-appropriate method
                )
}
```

#### Read csv file from bz2 archive:
```{r extract}
storm <- read.csv(bzfile("StormData.csv.bz2"), stringsAsFactors = F)
```

#### Data transformations: 
Transform data table to a tibble:
```{r tbl}
library(dplyr)
storm <- as.tbl(storm)
```

Leave only variables related to type of disaster, time of event, state where catastrophe happened, population health (fatalities, injuries) and economic damage (property, crop):
```{r subset_data}
storm_subset <- storm %>% select(EVTYPE, BGN_DATE, STATE, FATALITIES, INJURIES, PROPDMG, CROPDMG)
```

Change variables to appropriate formats:
```{r var_types}
# Type of the disaster
storm_subset$EVTYPE <- as.factor(storm_subset$EVTYPE)

# Date when event happened
    # used 'CST' timezone, because it is most common in the dataset 
    # check it: storm$TIME_ZONE %>% table()  
library(lubridate, verbose = F)
storm_subset$BGN_DATE <- storm_subset$BGN_DATE %>% 
    strptime(format = "%m/%d/%Y %H:%M:%S", tz = "CST") %>%
    as.POSIXct() %>% 
    year()  

# State where the event happened
storm_subset$STATE <- as.factor(storm_subset$STATE)
```

Change variable names:
```{r var_names}
names(storm_subset) <-  tolower(names(storm_subset))
storm_subset <-  storm_subset %>% 
    rename(event_type = evtype, year = bgn_date, property_damage = propdmg, crop_damage = cropdmg)
```  

Take a look at the current tibble:
```{r first_look}
storm_subset
```

The data is not tidy. It is messy because column headers are values, not variable names.
To fix it let's create `damage_type` and `damage` variables using `gather()` from `tidyr`:
```{r tidy}
library(tidyr)
storm_tidy <- storm_subset %>% 
    gather(key = damage_type, value = damage, -(event_type:state))
```
Let's take a look at tidy storm dataset:
```{r tidy_look}
storm_tidy
```



Split tidy tibble into two - one related to health damages (fatalities, injuries), another related to economic damages (property and crop damage):
```{r two_tibbles}
storm_health <- storm_tidy %>% filter(damage_type %in% c("fatalities", "injuries"))
storm_economic <- storm_tidy %>% filter(damage_type %in% c("property_damage", "crop_damage"))
```
Let's take a look at those tables:
```{r tidy_tibbles_look}
head(storm_health, 3)
head(storm_economic, 3)
```
Need to change `damage_type` in both tibbles to factor for further analysis:
```{r factor_damage_type}
storm_health$damage_type <- as.factor(storm_health$damage_type)
storm_economic$damage_type <- as.factor(storm_economic$damage_type)
```


## 3. Results
First, let's answer this question:  

> 1. Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?  

Looking at top 5 most damaging events related to health:
```{r top5_health_damage}
storm_health %>%
    group_by(event_type) %>%
    summarise(total_damage = sum(damage)) %>%
    arrange(desc(total_damage)) %>%
    head(5)
```

Let's build plot to compare top 5 most harmful events related to fatalities and injuries:
```{r top5_health_plot}
library(ggplot2)
library(ggthemes)

# Create a tibble with 5 most harful event related to fatalities
top5_fatalities <- storm_health %>% 
    filter(damage_type == "fatalities") %>% 
    group_by(event_type, damage_type) %>%
    summarise(total_damage = sum(damage)) %>%
    arrange(desc(total_damage)) %>%
    head(5)

# Create a tibble with 5 most harful event related to injuries
top5_injuries <- storm_health %>% 
    filter(damage_type == "injuries") %>% 
    group_by(event_type, damage_type) %>%
    summarise(total_damage = sum(damage)) %>%
    arrange(desc(total_damage)) %>%
    head(5)

# Combine two tibbles into one
top5_health <- bind_rows(top5_injuries, top5_fatalities)

# Reorder levels of damage_type, to display injuries first
top5_health$damage_type <-  factor(top5_health$damage_type, levels = c("injuries", "fatalities"))

# Plot stack bar plot
top5_health %>% 
    ggplot(aes(x = damage_type, y = total_damage, fill=event_type)) +
        geom_col() +
        scale_y_continuous(limits = c(0, 120000), breaks = seq(0, 120000, 20000)) +
        scale_fill_brewer(palette="Accent") +
        labs(x = "Type of health related damage",
             y = "Total damage (casualties)",
             fill = "Type of disaster") +
        ggtitle("Top 5 most harmful population health events in US: Injuries vs Fatalities", subtitle = "Years 1952 - 2011") +
        ggplot2::theme_classic()
```

And the second question: 

> 2. Across the United States, which types of events have the greatest economic consequences?   

Top 5 most damaging events related to economics:
```{r top5_economics_damage}
storm_economic %>%
    group_by(event_type) %>%
    summarise(total_damage = sum(damage)) %>%
    arrange(desc(total_damage)) %>%
    head(5)
```

Let's build plots to compare top 5 most harmful events related to property and health damages:
```{r top5_economics_plot}
library(ggplot2)
library(ggthemes)

# Create a tibble with 5 most harful event related to property damage
top5_prop_damage <- storm_economic %>% 
    filter(damage_type == "property_damage") %>% 
    group_by(event_type, damage_type) %>%
    summarise(total_damage = sum(damage)) %>%
    arrange(desc(total_damage)) %>%
    head(5)

# Create a tibble with 5 most harful event related to crop damage
top5_crop_damage <- storm_economic %>% 
    filter(damage_type == "crop_damage") %>% 
    group_by(event_type, damage_type) %>%
    summarise(total_damage = sum(damage)) %>%
    arrange(desc(total_damage)) %>%
    head(5)

# Combine two tibbles into one
top5_economics <- bind_rows(top5_prop_damage, top5_crop_damage)

# Reorder levels of damage_type, to display property damage first
top5_economics$damage_type <-  factor(top5_economics$damage_type, levels = c("property_damage", "crop_damage"))

# Turn off scientific notation
options(scipen=999)

# Plot stack bar plot
top5_economics %>% 
    ggplot(aes(x = damage_type, y = total_damage, fill=event_type)) +
        geom_col() +
        scale_y_continuous(limits = c(0, 8000000), breaks = seq(0, 8000000, 1000000)) +
        scale_fill_brewer(palette="Accent") +
        labs(x = "Type of economic related damage",
             y = "Total damage",
             fill = "Type of disaster") +
        ggtitle("Top 5 most economicly damaging events in US: Property vs Crop", subtitle = "Years 1952 - 2011") +
        ggplot2::theme_classic()
```
