---
title: "Storm Database analysis"
output: html_notebook
---



## 1. Synopsis

## 2. Data Processing  

#### Set global options for knitr:
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


#### Display when and on what system analysis was performed:
```{r date}
dateDownloaded <- date()
dateDownloaded
```

```{r system_info}
system_info <- sessionInfo()
system_info
```

#### Download storm dataset and extract related csv file:  
```{r download}
dlMethod <- "curl" # sets default for OSX / Linux
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(substr(Sys.getenv("OS"),1,7) == "Windows") dlMethod <- "wininet"

if(!file.exists("StormData.csv.bz2")) {
  download.file(url,
                "StormData.csv.bz2",  # stores file in R working directory
                method=dlMethod # use OS-appropriate method
                )
}
```

#### Read csv file from bz2 archive:
```{r extract}
storm <- read.csv(bzfile("StormData.csv.bz2"))
```

## 3. Results


Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*. 



Add a new chunk  *Ctrl+Alt+I*.

*Preview* button or press *Ctrl+Shift+K* to preview the HTML file.

