---
title: "Storm Database analysis"
output:
  html_notebook:
    highlight: tango
    theme: united
    toc: yes
---



## 1. Synopsis

## 2. Data Processing  

**Set global options for knitr:**
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


**Display when and on what system analysis was performed:**  
```{r date}
dateDownloaded <- date()
dateDownloaded
```

```{r system_info}
system_info <- sessionInfo()
system_info
```

**Download storm dataset and archive:**  
```{r download}
dlMethod <- "curl" # sets default for OSX / Linux
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(substr(Sys.getenv("OS"),1,7) == "Windows") dlMethod <- "wininet"

if(!file.exists("StormData.csv.bz2")) {
  download.file(url,
                "StormData.csv.bz2",  # stores file in R working directory
                method=dlMethod # use OS-appropriate method
                )
}
```

**Read csv file from bz2 archive:**
```{r extract}
storm <- read.csv(bzfile("StormData.csv.bz2"), stringsAsFactors = F)
```

**Data transformations:** 
Transform data table to a tible:
```{r tbl}
library(dplyr)
storm <- as.tbl(storm)
```

Leave only variables related type of disaster, time of an event, state where catastrophy happened, population health (fatalities, injuries) and economic damage (property, crop):
```{r subset_data}
storm_subset <- storm %>% select(EVTYPE, BGN_DATE, STATE, FATALITIES, INJURIES, PROPDMG, CROPDMG)
```

Change variables to appropriate formats:
```{r var_types}
# Type of the disaster
storm_subset$EVTYPE <- as.factor(storm_subset$EVTYPE)

# Date when event happened
    # used 'CST' timezone, because it is most common in the dataset 
    # check it: storm$TIME_ZONE %>% table()
library(lubridate, verbose = F)
storm_subset$BGN_DATE <- storm_subset$BGN_DATE %>% 
    strptime(format = "%m/%d/%Y %H:%M:%S", tz = "CST") %>%
    as.POSIXct() %>% 
    year()

# State where the event happened
storm_subset$STATE <- as.factor(storm_subset$STATE)
```


## 3. Results


Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*. 



Add a new chunk  *Ctrl+Alt+I*.

*Preview* button or press *Ctrl+Shift+K* to preview the HTML file.

